Understanding Implications of Dataset Choice for Feature Effect Estimation: A Simulation-Based Investigation through Error Decomposition

### 2025-02-15 paper/paper.tex
paper: add conclusion chapter and abstract

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{ragged2e}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{mwe}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

@@

\begin{abstract}
    The abstract should briefly summarize the contents of the paper in
    150--250 words.

    \keywords{Explainable AI  \and Feature effects \and Partial dependence plot \and Accumulated local effects}
\end{abstract}

\section{Introduction}
Most Machine Learning (ML) models can be considered black boxes --- opaque
systems that intrinsically do not allow insight into their internal reasoning,
making it often impossible to explain their decisions. However, this can be
problematic in many critical domains and applications, such as the healthcare,
legal, or finance sectors, where decisions must be transparent and
accountable~\cite{adadi_peeking_2018}.

Interpretability in ML is crucial to enhance
trust~\cite{ribeiro_why_2016,teach_analysis_1981}, address potential
biases~\cite{guidotti_survey_2019}, fairness and ethical concerns
~\cite{lipton_mythos_2018}, and ensure compliance with regulations such as the
EU's General Data Protection Regulation (GDPR)~\cite{gdpr2016} and AI
Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
/ Interpretable ML has emerged~\cite{adadi_peeking_2018}. In the last years, a
wide variety of methods has been developed, including model-specific and
model-agnostic approaches ranging from local feature attributions to global
feature importances and effects\footnote{For an overview of Explainable AI
    methods, see
    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.}.

Due to the severity of many applications, it is crucial to utilize these
explainability methods correctly to avoid misleading or incorrect conclusions.
In general, there are many pitfalls to be aware of~\cite{molnar_general_2022},
including whether to compute explanations in-sample, i.e.\ on training data, or
out-of-sample, which we refer to as validation data in the following. Existing
works have studied the implications of this choice for many methods, including
studies on \textit{Permutation Feature Importance (PFI) } (e.g.,
in~\cite{molnar_general_2022}), \textit{Mean Decrease in Impurity (MDI)} of
Random Forests~\cite{loecher_debiasing_2022}, or \textit{SHAP}
values~\cite{loecher_debiasing_2024}.

However, to the best of our knowledge, there exist no such systematic studies
for feature effect methods like \textit{Partial Dependence Plots
    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
    (ALE)}~\cite{apley_visualizing_2020}. Current literature
(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022})
predominantly uses training data without explicit justification, while
practitioners also advocate for using unseen validation data (for details, see
\textsc{Section~\ref{sec:related-works}}). Factors that may influence this
choice include potential biases arising from overfitting, the dataset size, and
computational constraints. Although PDP and ALE are not based on generalization
error like PFI, it is not studied if and how they are affected by overfitting.
In addition, larger datasets may improve the feature effect estimates, but also
increase computation times.\\

\noindent In this paper, we aim to answer the largely unaddressed, fundamental
methodological question of whether to use training or validation data to
compute feature effects. We perform an empirical simulation study, estimating
PDP and ALE on training data, validation data, and in a cross-validated manner
for different models and dataset. We decompose the error of PDP and ALE and
compare the error components to understand the implications of the data choice.
Our main contributions can be summarized as follows:

@@

\noindent These contributions have several important implications: Firstly, our empirically
grounded recommendations enable practitioners and researchers to make informed
decisions about which data to choose for feature effect computation, helping to
understand potential implications of their choices. Additionally, our framework
for evaluating feature effects and systematic collection of test functions
provide a foundation for future research in Interpretable ML, with the latter
specifically facilitating test function choice for simulation studies.\\

@@

\paragraph{Feature Effect Methods.}
Current literature offers various methods for analyzing feature effects in ML. % chktex 13
One of the most popular methods is the \textit{Partial Dependence Plot
    (PDP)}~\cite{friedman_greedy_2001}, which describes the marginal effect of one
or two features on the model prediction. The PDP assumes that the features of
interest are independent of the remaining features. When this assumption is
violated, the method may produce unrealistic data points outside the underlying
joint distribution of the data. This extrapolation issue can cause misleading
interpretations~\cite{molnar_interpretable_2022,molnar_general_2022}.

@@

Alternative approaches include \textit{functional ANOVA (fANOVA)}, which
decomposes feature effects into main and interaction
effects~\cite{hooker_discovering_2004}. Recent extensions have further
addressed limitations of these classical methods. For example, \textit{Robust
    and Heterogeneity-aware ALE (RHALE)}~\cite{gkolemis_rhale_2023} adds
heterogeneity quantification to ALE and improving robustness, or
\textit{Accumulated Total Derivative Effect (ATDEV)} plots can be decomposed
into ALE and \textit{Accumulated Cross Effects (ACE)}~\cite{liu_model_2018}.

@@

\paragraph{Feature Effect Error and Uncertainty.}
Multiple works have approached measuring uncertainty in feature effects. For
probabilistic ML models, Moosbauer et al.\cite{moosbauer_explaining_2021}
derived model-specific confidence bands for PDPs, while applied studies have
proposed bootstrap-based
approaches\cite{esselman_landscape_2015,grange_using_2019}. While the former
are not model-agnostic, the latter often capture only the variance introduced
by Monte Carlo approximation. Nonetheless, the model variance is another source
of uncertainty in PDPs and ALEs, and accounting for it requires multiple model
fits~\cite{apley_visualizing_2020,molnar_general_2022}.

@@

We give the results of the main experiment error decomposition for the SimpleNormalCorrelated
dataset in \textbf{Fig.\@~\ref{fig:pdp-results-snc}} for PDP and \textbf{Fig.\@~\ref{fig:ale-results-snc}}
for ALE. The full results for all datasets, models, and features can be found in  % chktex 13
\textsc{Appendix~\ref{app:additional-results}}.

Generally, we observe that the feature effect error is in the most cases highest when
the effect is estimated on validation data. For ALE, this observation is more pronounced
than for PDP. This increase in error is mainly due to an increase in variance (see e.g.,  % chktex 13
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or \textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}).
For many models, the feature effect error is lowest when estimated in a cross-validated manner.
This is slightly more pronounced for overfitting models (OF) than for optimally tuned models (OT),
and is mainly due to a decrease in variance (see e.g., \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}}
and \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes of the model variance
and the Monte Carlo variance, both could be reduced by cross-validation. Monte Carlo variance
may decrease as we have slightly more samples overall than in the training set. Conversely,
the model variance may decrease as it is averaged out through the cross-validation procedure,
which may especially improve overfitting models that have a higher model variance.
We further investigate these hypotheses in \textsc{Section~\ref{sec:results-variance-decomposition}}.

@@

Further results on Friedman1 and Feynman I.29.16, as well as for the dummy features, largely confirm
the described findings. For details, we refer to \textsc{Appendix~\ref{app:additional-results}}. We only
observe one case where the error of ALE is substantially larger when estimated on training data
due to a high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset, which we further
look into in \textsc{Appendix~\ref{app:main-study-further-investigation}}. Also note that the we observe
no substantial differences between scenarios with correlated features (SimpleNormalCorrelated) and
independent features (Friedman1, Feynman I.29.16).

@@

We use the variance results from the main experiment and the Monte Carlo variance estimated in the ablation study
to show the decomposition of the feature effect variance into model variance and Monte Carlo variance. The results
for the Friedman1 dataset are given in \textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
\textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. The full results including the other datasets are  % chktex 13
provided in \textsc{Appendix~\ref{app:additional-variance-decomposition}}.

@@

In addition, we observe that while the model variance is naturally nearly the same for training and validation data,
it is also reduced in many cases when estimating the feature effects in a
cross-validated manner. This is especially pronounced for the overfitting models (OF) and is best visible for PDP
(\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line with our previous hypothesis that
reduces both model variance by averaging over different models as well as Monte Carlo variance by averaging over
different samples. This is especially beneficial for overfitting models, which have a higher model variance.

@@

While the variance due to MC integration is only a minor component of the total variance for PDP, it is the
dominant component for ALE. Only for feature effect estimation via cross-validation, the Monte Carlo variance  % chktex 13
is again reduced to a negligible level. This suggests that ALE is more sensitive to the dataset size than PDP,
which we further investigate in \textsc{Section~\ref{sec:results-dataset-size}}. We also suspect a certain
amount of inaccuracy in the Monte Carlo variance estimation for ALE, which (a) leads to differences in the
model variance between training and validation set and (b) causes the model variance to be negative in some cases
(see \textbf{Fig.\@\ref{fig:ale-variance-decomp}}).
Naturally, this is not possible, and can most likely be attributed to numerical inaccuracies in the MC variance estimation.

@@

For both PDP and ALE, we find that the Monte Carlo variance, as expected, reduces with larger dataset size.
This becomes clear both when comparing training and validation data and when comparing the two dataset sizes
(1000 and 8000 samples). To analyze this not only for these few dataset sizes, we dedicate a separate
ablation study to investigate this further in the next section.

@@

We show the effect of the dataset size on the Monte Carlo variance for the features of the Friedman1 dataset
for PDP and ALE in \textbf{Fig.\@~\ref{fig:mc-variance}}. The full results are provided in \textsc{Appendix~\ref{app:additional-mc-results}}.
Naturally, we observe that the Monte Carlo variance decreases with increasing dataset size (except for $x_3$ and $x_4$ for PDP,
where the variance is generally very small). We also see differences between the features: for more complex features like $x_1$ and $x_2$,
which include interactions, the Monte Carlo variance is generally higher compared to simpler features like $x_3$ and $x_4$.
Nonetheless, we observe an interesting difference between PDP and ALE. While they converge to a similar Monte Carlo variance for  % chktex 13
large dataset sizes, the initial Monte Carlo variance for very small sample sizes is substantially higher for ALE than for PDP.  % chktex 13
This demonstrates that ALE is more sensitive to the dataset size than PDP. This is most likely due to the fact that ALE  % chktex 13
estimates effects per interval and thus requires more samples to estimate the effect accurately compared to PDP, which
uses the full dataset to estimate the effect. Also note that this analysis shows that the Monte Carlo variance introduced
by estimating the ground truth feature effects in the main study is for most cases negligible compared to the other error sources.

@@

@@

\begin{credits}
    \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is used for
    general acknowledgments, for example: This study was funded by X (grant number
    Y).

    \subsubsection{\discintname}
    The authors have no competing interests to declare that are relevant to the
    content of this article.
\end{credits}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{ragged2e}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{mwe}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

@@

\begin{abstract}
    As Machine Learning models are increasingly employed in critical domains
    like healthcare and finance, ensuring accurate model explanations by
    Interpretable Machine Learning methods becomes crucial to avoid misguided
    decisions. The choice of dataset for estimating feature effects --- whether
    to use training data, validation data, or even employ cross-validation ---
    is one important fundamental question in this context that remains largely
    unexplored.
    Through a comprehensive simulation study, we investigate this question for
    Partial Dependence Plots (PDP) and Accumulated Local Effects (ALE), decomposing
    their error into bias and variance components. We extend previous work by
    providing a theoretical framework for error decomposition and contribute
    a systematic overview of test functions for simulation studies in Interpretable ML.  % chktex 13
    Our simulation study results validate the common practice of using training
    data, showing no evidence against it. We find that larger datasets for estimation
    are generally preferable due to reduced Monte Carlo variance, with ALE being
    particularly sensitive to dataset size. Cross-validation emerges as a promising
    approach, reducing both model and Monte Carlo variance, which is especially
    beneficial for overfitting models.
    Our findings enable practitioners to make informed decisions about dataset
    choice for feature effect estimation while establishing a foundation for
    future research in this area.\footnote{Our entire implementation and results
    are available at \url{https://github.com/timo282/current-research-feature-effects}.}

    \keywords{Interpretable Machine Learning  \and Feature effects \and Partial dependence plot \and Accumulated local effects}
\end{abstract}

\section{Introduction}
Most Machine Learning (ML) models can be considered black boxes --- opaque
systems that intrinsically do not allow insight into their internal reasoning,
making it often impossible to explain their decisions. This can be
problematic in many critical domains and applications, such as healthcare,
legal, or finance, where decisions must be transparent and
accountable~\cite{adadi_peeking_2018}.

The need for interpretability in ML stems from multiple factors: building trust in
model predictions~\cite{ribeiro_why_2016,teach_analysis_1981},  identifying and mitigating
potential biases~\cite{guidotti_survey_2019}, addressing fairness and ethical concerns
~\cite{lipton_mythos_2018}, and ensuring compliance with regulations like the
EU's General Data Protection Regulation (GDPR)~\cite{gdpr2016} and AI
Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
/ Interpretable ML has emerged~\cite{adadi_peeking_2018}, develping a
wide variety of methods to explain ML models. This includes both model-specific and
model-agnostic approaches ranging from local feature attributions to global
feature importances and effects\footnote{For comprehensive overviews of Explainable AI
    methods, see
    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.}.

Due to the severity of many applications, it is crucial to utilize these
interpretability methods correctly to avoid misleading or incorrect conclusions.
In general, there are many pitfalls to be aware of~\cite{molnar_general_2022},
including the question whether to compute explanations in-sample, i.e.\ on training data, or
out-of-sample, which we refer to as validation data in the following. Existing
works have studied the implications of this choice for many methods, including
studies on \textit{Permutation Feature Importance (PFI) }~\cite{molnar_general_2022},
\textit{Mean Decrease in Impurity (MDI)} of
Random Forests~\cite{loecher_debiasing_2022}, or \textit{SHAP}
values~\cite{loecher_debiasing_2024}.

However, to the best of our knowledge, there exist no such systematic studies
for feature effect methods like \textit{Partial Dependence Plots
    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
    (ALE)}~\cite{apley_visualizing_2020}. Current literature
(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022})
predominantly uses training data without explicit justification, while
practitioners also advocate for using unseen validation data (for details, see
\textsc{Section~\ref{sec:related-works}}). Factors that may influence this
choice include potential biases arising from overfitting, the dataset size, and
computational constraints. Although PDP and ALE are not based on generalization
error like PFI, it is not studied if and how they are affected by overfitting
when estimated on training data. In addition, larger datasets may improve the
feature effect estimates, but also increase computation times.\\

\noindent In this paper, we aim to answer this largely unaddressed, fundamental
methodological question of whether to use training or validation data to
compute feature effects. We perform an empirical simulation study, estimating
PDP and ALE on training data, validation data, and in a cross-validated manner
for different models and dataset. We decompose the error of PDP and ALE and
compare the error components to understand the implications of the data choice.
Our main contributions can be summarized as follows:

@@

\noindent These contributions have several important implications: First, our empirically
grounded recommendations enable practitioners and researchers to make informed
decisions about which data to choose for feature effect estimation, helping to
understand potential implications of their choices. Additionally, our framework
for evaluating feature effects and systematic collection of test functions
provide a foundation for future research in Interpretable ML, with the latter
specifically facilitating test function choice for simulation studies.\\

@@

\paragraph{Feature Effect Methods.}
Current literature offers various methods for analyzing feature effects in ML.  % chktex 13
One of the most popular methods is the \textit{Partial Dependence Plot
    (PDP)}~\cite{friedman_greedy_2001}, which describes the marginal effect of one
or two features on the model prediction. The PDP assumes that the features of
interest are independent of the remaining features. When this assumption is
violated, the method may produce unrealistic data points outside the underlying
joint data distribution of the data, which can cause misleading
interpretations (``extrapolation issue'')~\cite{molnar_interpretable_2022,molnar_general_2022}.

@@

Alternative approaches include \textit{functional ANOVA (fANOVA)}, which
decomposes feature effects into main and interaction
effects~\cite{hooker_discovering_2004}. Recent extensions have further
addressed limitations of the methods above: \textit{Robust
    and Heterogeneity-aware ALE (RHALE)}~\cite{gkolemis_rhale_2023} adds
heterogeneity quantification to ALE and improves robustness.
\textit{Accumulated Total Derivative Effect (ATDEV)} plots can be decomposed
into ALE and \textit{Accumulated Cross Effects (ACE)}~\cite{liu_model_2018}.

@@

\paragraph{Feature Effect Error and Uncertainty.}
Multiple works have approached measuring uncertainty in feature effects. For
probabilistic ML models, Moosbauer et al.\cite{moosbauer_explaining_2021}
derived model-specific confidence bands for PDPs, while applied studies have
proposed bootstrap-based
approaches~\cite{esselman_landscape_2015,grange_using_2019}. While the former
are not model-agnostic, the latter often capture only the variance introduced
by Monte Carlo approximation. However, the model variance is another source
of uncertainty in PDPs and ALEs, and accounting for it requires multiple model
fits~\cite{apley_visualizing_2020,molnar_general_2022}.

@@

We give the results of the main experiment error decomposition for the
SimpleNormalCorrelated dataset in \textbf{Fig.\@~\ref{fig:pdp-results-snc}} for
PDP and \textbf{Fig.\@~\ref{fig:ale-results-snc}} for ALE. The full results for  % chktex 13
all datasets, models, and features can be found in
\textsc{Appendix~\ref{app:additional-results}}.

Generally, we observe that the feature effect error is in the most cases
highest when the effect is estimated on validation data. For ALE, this
observation is more pronounced than for PDP. This increase in error is mainly  % chktex 13
due to an increase in variance (see e.g.,
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or
\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}). For many models, the
feature effect error is lowest when estimated in a cross-validated manner. This
is slightly more pronounced for overfitting models (OF) than for optimally
tuned models (OT), and is mainly due to a decrease in variance (see e.g.,
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}} and
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes
of the model variance and the Monte Carlo variance, both could be reduced by
cross-validation. Monte Carlo variance may decrease as we have slightly more
samples overall compared to the training set. Conversely, the model variance
may decrease as it is averaged out through the cross-validation procedure,
which may especially improve overfitting models that have a higher model
variance. We further investigate these hypotheses in
\textsc{Section~\ref{sec:results-variance-decomposition}}.

@@

Overall, we find no evidence against the currently dominant practice of
estimating feature effects on training data. Further results on Friedman1 and
Feynman I.29.16, as well as for the dummy features, largely confirm our
described findings. For details, we refer to
\textsc{Appendix~\ref{app:additional-results}}. This also means that we observe
no substantial differences between scenarios with correlated features
(SimpleNormalCorrelated) and independent features (Friedman1, Feynman I.29.16).
Similarly, the more real-world relevant Feynman I.29.16 dataset shows results
in line with the other datasets. Note that we only observe one case where the
error of ALE is substantially larger when estimated on training data due to a
high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset,
which we further look into in
\textsc{Appendix~\ref{app:main-study-further-investigation}}.

@@

We use the variance results from the main experiment and the Monte Carlo
variance estimated in the ablation study to show the decomposition of the
feature effect variance into model variance and Monte Carlo variance. The
results for the Friedman1 dataset are given in
\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
\textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. Note that we now  % chktex 13
consider the Friedman1 dataset as it provides a wider variety of different
feature effects. The full results including the other datasets are provided in
\textsc{Appendix~\ref{app:additional-variance-decomposition}}.

@@

In addition, we observe that the model variance is also reduced in many cases when
estimating the feature effects in a cross-validated manner compared to an estimation
on training or validation data, where model variance is (roughly) the same. The observed
reduction through cross-validation is especially
pronounced for the overfitting models (OF) and best visible for PDP
(\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line
with our previous hypothesis that cross-validation reduces both model variance by averaging over
different models as well as Monte Carlo variance by averaging over different
samples. This is especially beneficial for overfitting models, which have a
higher model variance.

@@

While the variance due to MC integration is only a minor component of the total
variance for PDP, it is the dominant component for ALE. Only for feature effect  % chktex 13
estimation via cross-validation, the Monte Carlo variance is again reduced to a
negligible level. This suggests that ALE is more sensitive to the dataset size
than PDP, which we further investigate in
\textsc{Section~\ref{sec:results-dataset-size}}. We also suspect a certain
amount of inaccuracy in the Monte Carlo variance estimation for ALE, which (a)
leads to differences in the model variance between training and validation set
and (b) causes the model variance to become negative in some cases (see
\textbf{Fig.\@\ref{fig:ale-variance-decomp}}). Naturally, this is not possible
and can most likely be attributed to numerical inaccuracies in the MC variance
estimation.

@@

For both PDP and ALE, we find that the Monte Carlo variance, as expected,
reduces with larger dataset size. This becomes clear both when comparing
training and validation data and when comparing the two dataset sizes (1000 and
8000 samples). To analyze this not only for those few dataset sizes, we
dedicate a separate ablation study to investigate this further in the next
section.

@@

We show the effect of the dataset size on the Monte Carlo variance for the
features of the Friedman1 dataset for PDP and ALE in
\textbf{Fig.\@~\ref{fig:mc-variance}}. The full results are provided in
\textsc{Appendix~\ref{app:additional-mc-results}}. As expected, we observe that
the Monte Carlo variance decreases with increasing dataset size (except for
$x_3$ and $x_4$ for PDP, where the variance is generally very small). We
find differences between the features: for more complex features like $x_1$ and
$x_2$, which include interactions, the Monte Carlo variance is generally higher
compared to simpler features like $x_3$ and $x_4$. Nonetheless, we observe an
interesting difference between PDP and ALE. While they converge to a similar  % chktex 13
Monte Carlo variance for large dataset sizes, the initial Monte Carlo variance
for very small sample sizes is substantially higher for ALE than for PDP. This  % chktex 13
demonstrates that ALE is more sensitive to the dataset size than PDP. This is  % chktex 13
most likely due to the fact that ALE estimates effects per interval and thus
requires more samples to estimate the effect accurately compared to PDP, which
uses the full dataset to estimate the effect. Note that this analysis
also proves that the Monte Carlo variance introduced by estimating the ground truth
feature effects in the main study is for most cases negligible compared to the
other error sources.

@@

Summarizing our main contributions, we (1) empirically investigated the
impact of dataset choice for feature effect estimation. We (2) based our study on
a theoretical framework that decomposes the feature effect error into different
components and introduces estimators for each of them. We (3) conducted a
review of test functions for simulation studies and build the choices for
our study on these.

Our comprehensive simulation study provided several key insights into
the choice of dataset for feature effect estimation. First and foremost,
we find no general evidence that speaks against the current predominant practice
of using training data for computing feature effects. This is an important validation
of existing approaches in the literature.

More specifically, our results demonstrated that feature effect errors are
generally lower when more data is available for the estimation, which is usually
the case for the training set. This is primarily due to reduced Monte Carlo variance,
as shown in our ablation studies. The effect is particularly pronounced for ALE plots,
which exhibit higher sensitivity to dataset size compared to PDPs, especially for smaller samples,
due to the interval-based estimation. This suggests that for ALE plots, it is even more
important to use a sufficently large dataset for accurate estimation. Based on these
findings, we generally recommend using the largest available dataset for feature effect
estimation, as long as this is computationally feasible.

An interesting finding emerged regarding cross-validation for feature effect estimation.
Our main study suggests that cross-validation is preferable in many scenarios,
leading to lower overall errors. The ablation study reveals that this improvement stems from two distinct mechanisms:
(1) reduced Monte Carlo variance, and (2) reduced model variance through the averaging of estimates from different models.
This dual benefit is particularly advantageous for overfitting models, which typically suffer
from higher variance. Therefore, we recommend using cross-validation for feature effect estimation
when overfitting is a serious concern (e.g., due to a large difference between training
and validation performance). Estimating feature effects for a single model and dataset
or using cross-validation also becomes a trade-off between computation time and accurate
feature effect estimation, as cross-validation comes with computational overhead depending
on the number of folds.

We also observed that the correlation structure between features appears to have limited impact on our findings
and are largely consistent across our set of different, diverse datasets.
This suggests that our recommendations are robust across different data scenarios,
including those with complex feature interactions, correlations, 
and real-world inspired relationships as in the Feynman dataset.

@@

While our study provides valuable insights, several limitations and opportunities
for future research remain. One methodological limitation lies in our choice of using
theoretical quantiles for the grid points in feature effect estimation to make feature
effects comparable. However, this can lead to empty or sparsely populated intervals, particularly for smaller
datasets or in regions of low data density. Moreover, this approach has limited real-world
applicability, as we usually do not now the theoretical but only the empirical quantiles
(which should be the same in expectation).

Furthermore, the observed instabilities in variance estimation for the Monte Carlo variance,
particularly of ALE plots, indicate room for methodological improvements. Future research may
focus on developing more robust estimation procedures for decomposing variance components,
before applying these estimators to address other research questions.

Another limitation of our study lies in the pre-selection of (optimal) model hyperparameters.
In practice, these parameters need to be tuned typically using the same dataset as for feature
effect estimation. This introduces a potential post-selection bias requiring for more sophisticated
methods, especially when estimating the feature effects in a cross-validated manner, such as
the use of nested cross-validation.

Further, it is important to note that our study primarily adopts a ``true to the data'' perspective
when defining feature effect errors, focusing on explaining the underlying data-generating process
rather than the model's internal representation. While this choice aligns with many practical
applications, there are also use cases for which ``true to the model'' interpretations might be 
preferred, which is not considered in our analysis.
% also: data choice only with regard to feature effect error, not uncertainty estimation

Future research could also investigate the impact of data shifts between training and validation sets.
While our focused considered overfitting models, future work could explore the impact of data shifts
between training and validation sets and their impact on feature effect estimation, analyzing the
robustness of the estimation under distribution shifts.

Moreover, while our cross-validation strategy showed promising improvements over
conventional estimation on a single dataset, this approach could be enhanced further.
Future research could explore strategies that utilize more data per fold, such as using the
training set in each fold for feature effect estimation (and use the validation part only for
perfomance estimation) or combining training and validation splits. Alternatively, bootstrapping
strategies similar to the approaches pursued by Molnar et al.~\cite{molnar_relating_2023} could be
explored.

Following the observations of Loecher~\cite{loecher_debiasing_2024} that hat MDI and SHAP show
a bias toward high-entropy variables for Random Forests, future work might also investigate if and how
feature entropy and cardinality influence feature effect estimation.

While we aimed to incorporate real-world relevance through physics-inspired datasets, further
work is needed to fully capture the complexity of practical applications. In practice, datasets
have features with varying and also just partial effect strengths, different noise levels,
higher-order interactions, and diverse marginal distributions. Future work could focus on
developing more realistic datasets capturing these aspects, to further bridge the gap between
research, such as simulation studies, and practical applications.

These limitations and future directions highlight the rich opportunities for further research
in this area to obtain practically relevant insights.

\begin{credits}
    % \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is used for
    % general acknowledgments, for example: This study was funded by X (grant number
    % Y).
    \subsubsection{\discintname}
    The authors have no competing interests to declare that are relevant to the
    content of this article.
\end{credits}
